\s12

.ce
Rochester Institute of Technology
.ce
School of Computer Science and Applied Technology


\s15
.ce
.B "ClassC/Elaine:"

.ce
.I "A Multiple Inheritance Object Oriented C Language"
\s12


.ce
by
.ce
Paul Kirkaas


A thesis submitted to the faculty of the School of Computer Science and
Technology in partial fulfillment of the requirements for the degree of
Master of Science in Computer Science.

Approved by:


__________________________________________________________
.br
Dr James Heliotis



__________________________________________________________
.br
Dr Andrew Kitchen



__________________________________________________________
.br
Ex Officio Committee Member
.bp













.ce
.I "ClassC/Elaine:"
.ce
.I "A Multiple Inheritance Object Oriented C Language"


.ce
by Paul Kirkaas









I, Paul Kirkaas, hereby grant permission to the Wallace Memorial Library of
RIT to reproduce my thesis in whole or in part provided the source is
credited.  Any reproduction will not be commercial use or profit.








________________________________________

Paul Kirkaas

12 September 1988
.bp




.ce
.I ABSTRACT

.ls 1
Object oriented programming is a way of abstracting information and
operations to make programming more efficient and reliable.
.I C
is a non object oriented programming language that has become a de facto
language standard in academic and industrial applications because of its
power and flexibility.  ClassC is an attempt to add object orientation
on the existing framework that C provides.
Specifically, ClassC adds the new data type
.I object
, and the new aggregate declaration,
.I class.
.br
        ClassC differs from other C-based object oriented languages such as C++
Objective-C in three points:
  1) ClassC provides true multiple inheritance, which is very commonly
discussed in the description of object oriented languages, but very
rarely actually implemented.
  2) ClassC provides both strict type checking on objects
as well as typeless dynamic binding -- a variable of type
.I object
may be assigned any class instantiation.  Dynamic binding is
typical of interpreted
languages like lisp and Smalltalk, but rare in compiled languages like
.I C
and
.I Pascal,
etc.
  3) ClassC offers automatic garbage collection of memory no longer
referenced by any object variables.
        ClassC/Elaine is intended to be a flexible system that
allows programmers to
explore different qualities of object oriented programming, including strong
and weak type checking, and single and multiple inheritance. This thesis is
a description of the background, design, and implementation
of the ClassC language.
.ls 2
